---
title: "jacobson_cores"
output: github_document
date: "2023-09-29"
---

load Packages 
```{r}
library(tidyverse)
library(utils)
library(dplR)
library(reshape2) 
```

Core Data

Load Data - Tree Core Ring Widths 
```{r}
jack_cores <- read_delim("jack_corescsdoneTAB_C.txt") %>% column_to_rownames(var = "Year")
chris_cores <- read_delim("Chris_corescsTAB_C.txt") %>% column_to_rownames(var = "Year")
```


Data Cleaning 
```{r }
J_ids <- read.ids(jack_cores) %>% 
  rownames_to_column("core_name") %>% 
  mutate(core_name1 = str_sub(core_name, 1, -3)) %>% 
  group_by(core_name1) %>% 
  mutate(core1 = 1:n()) %>% 
  ungroup() %>% 
  select(core_name, core_name1, core1) %>% 
  column_to_rownames("core_name") %>% 
  rename(core = core1, tree = core_name1)
  colnames(jack_cores) <- make.names(colnames(jack_cores))
  
C_ids <- read.ids(chris_cores) %>% 
  rownames_to_column("core_name") %>% 
  mutate(core_name1 = str_sub(core_name, 1, -3)) %>% 
  group_by(core_name1) %>% 
  mutate(core1 = 1:n()) %>% 
  ungroup() %>% 
  select(core_name, core_name1, core1) %>% 
  column_to_rownames("core_name") %>% 
  rename(core = core1, tree = core_name1)
  colnames(chris_cores) <- make.names(colnames(chris_cores))
  C_ids <- na.omit(C_ids)
```

Tree Means 
```{r}
  
Jack_cores_rw <- treeMean(jack_cores, J_ids)
chris_cores_rw <- treeMean(chris_cores, C_ids)
```


calculate bai - This converts ring-width series (mm) to ring-area series (mm squared) (aka basal area increments) based on the diameter of the tree and the width of each ring moving towards the pith of the tree.
```{r}
jack_cores_bai <- bai.out(Jack_cores_rw)
chris_cores_bai <- bai.out(chris_cores_rw)
```

write to csv 
```{r, eval=FALSE}
#write.csv(jack_cores_bai, "D:/jack_cores/jack_cores_bai.csv", row.names = T)

#write.csv(Jack_cores_rw, "D:/jack_cores/jack_cores_rw.csv", row.names = T)
```

```{r}
bai_long <- jack_cores_bai %>% 
  rownames_to_column("Year") %>% 
  pivot_longer(!Year, names_to = "core_id", values_to = "measurement") %>% 
  arrange(core_id, Year) %>%
  separate(core_id, c("Forest", "Plot", "Tree"), "_") %>% 
  mutate(Year = as.numeric(Year))

TBbai_long <- chris_cores_bai %>% 
  rownames_to_column("Year") %>% 
  pivot_longer(!Year, names_to = "core_id", values_to = "measurement") %>% 
  arrange(core_id, Year) %>%
  separate(core_id, c("Forest", "Plot", "Tree"), "_") %>% 
  mutate(Year = as.numeric(Year))
```

```{r}
bai_CS <- bai_long %>% filter(Forest == "CS")

bai_TW51 <- bai_long %>% filter(Forest == "TW51")

bai_TB <- TBbai_long %>% filter(Forest == "TB")

bai_CL<- TBbai_long %>% filter(Forest == "CL")
```


```{r}
bai__avg_cs <- bai_CS %>%
  subset(select = -c(Forest, Plot, Tree)) %>% 
  group_by(Year) %>%
  summarise(average = mean(measurement, na.rm = TRUE))


ggplot(bai__avg_cs, aes(x = as.numeric(Year), y = average)) +
  geom_line(stat="identity",color="brown") +
  theme_minimal() +
  labs(title="Average BAI by Year for Craig Springs", x="Year",  y = expression("Average BAI"(mm^2)))+
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
bai__avg_TW51 <- bai_TW51 %>%
  subset(select = -c(Forest, Plot, Tree)) %>% 
  group_by(Year) %>%
  summarise(average = mean(measurement, na.rm = TRUE))


ggplot(bai__avg_TW51, aes(x = as.numeric(Year), y = average)) +
  geom_line(stat="identity",color="red") +
  theme_minimal() +
  labs(title="Average BAI by Year for Talking Warrior", x="Year", y = expression("Average BAI"(mm^2))) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
bai__avg_TB <- bai_TB %>%
  subset(select = -c(Forest, Plot, Tree)) %>% 
  group_by(Year) %>%
  summarise(average = mean(measurement, na.rm = TRUE))


p <-ggplot(bai__avg_TB, aes(x = as.numeric(Year), y = average)) +
  geom_line(stat="identity",color="maroon") + 
  theme_minimal() +
  labs(title="Average BAI by Year for Tombigbee", x="Year", y = expression("Average BAI" (mm^2)))+
  theme(plot.title = element_text(hjust = 0.5))
  p + scale_x_continuous(breaks = seq(min(bai__avg_TB$Year), max(bai__avg_TB$Year), by = 5), labels = seq(min(bai__avg_TB$Year), max(bai__avg_TB$Year), by = 5))
```
```{r}
bai__avg_CL <- bai_CL %>%
  subset(select = -c(Forest, Plot, Tree)) %>% 
  group_by(Year) %>%
  summarise(average = mean(measurement, na.rm = TRUE))

CL_avgplot<-ggplot(bai__avg_CL, aes(x = as.numeric(Year), y = average)) +
  geom_line(stat="identity",color="sienna") +
  theme_minimal() +
  labs(title="Average BAI by Year for Chocktaw Lake", x="Year", y = expression("Average BAI"(mm^2))) +
  theme(plot.title = element_text(hjust = 0.5))+
  coord_cartesian(xlim = c(1999, max(bai_CL$Year)))
print(CL_avgplot)
```




```{r}
bai__avg <- bai_long %>%
  subset(select = -c(Forest, Plot, Tree)) %>% 
  group_by(Year) %>%
  summarise(average = mean(measurement, na.rm = TRUE))


ggplot(bai__avg, aes(x = Year, y = average)) +
  geom_col() + 
  labs(title="Average BAI by Year", x="Year", y="Average BAI") +
  theme(plot.title = element_text(hjust = 0.5))

```


```{r}
bai__avg <- TBbai_long %>%
  subset(select = -c(Forest, Plot, Tree)) %>% 
  group_by(Year) %>%
  summarise(average = mean(measurement, na.rm = TRUE))


ggplot(bai__avg, aes(x = Year, y = average)) +
  geom_col() + 
  labs(title="Average BAI by Year", x="Year", y="Average BAI")

```
```{r}
# Filter the data for Craig Springs (CS)
bai_CS <- bai_long %>% filter(Forest == "CS")

# Create a single plot for all CS plots
combined_plot_CS <- ggplot(bai_CS, aes(x = as.numeric(Year), y = measurement, color = Plot)) +
  geom_line() +
  theme_minimal(base_size = 18) +
  labs(title = "Craig Springs", x = "Year", y = expression("Average BAI"(mm^2))) +
 theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none") +
  geom_vline(xintercept = 2020)
ggsave(combined_plot_CS, filename = "combined_plot_CS.png", dpi = 600, width = 12, height = 7, units = "in")

```
```{r}
# Filter the data for Talking Warrior (TW51)
bai_TW51 <- bai_long %>% filter(Forest == "TW51")

# Create a single plot for all TW51 plots
(combined_plot_TW51 <- ggplot(bai_TW51, aes(x = as.numeric(Year), y = measurement, color = Plot)) +
  geom_line() +
  theme_minimal(base_size = 18) +
  labs(title = "Talking Warrior", x = "Year", y = expression("Average BAI"(mm^2))) +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none") +
  geom_vline(xintercept = 2020))

ggsave(combined_plot_TW51, filename = "combined_plot_TW51.png", dpi = 600, width = 12, height = 7, units = "in")

# Display the combined plot for Talking Warrior
#print(combined_plot_TW51)
```


```{r}
bai_TB <- TBbai_long %>% filter(Forest == "TB")

(combined_plot_TB <- ggplot(bai_TB, aes(x = as.numeric(Year), y = measurement, color = Plot)) +
  geom_line() +
  theme_minimal(base_size = 18) +
  labs(title = "Tombigbee", x = "Year", y = expression("Average BAI"(mm^2)))+
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none") +
  geom_vline(xintercept = 2017))
  
ggsave(combined_plot_TB, filename = "combined_plot_TB.png", dpi = 600, width = 12, height = 7, units = "in")
 
```

```{r}
bai_CL <- TBbai_long %>% filter(Forest == "CL")
bai_CL <- bai_CL %>% mutate(Year = as.numeric(Year))

(combined_plot_CL <- ggplot(bai_CL, aes(x = as.numeric(Year), y = measurement, color = Plot)) +
  geom_line() +
  theme_minimal(base_size = 18) +
  labs(title = "Choctaw Lake", x = "Year", y = expression("Average BAI"(mm^2))) +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none") +
    coord_cartesian(xlim = c(2000, max(bai_CL$Year)))+
    geom_vline(xintercept = 2017))
    

  
ggsave(combined_plot_CL, filename = "combined_plot_CL.png", dpi = 600, width = 12, height = 7, units = "in")

  
```
```{r}
starr_burn <- 2020
fs_burn <- 2017

fire_pre_post_bai_CS <-
  bai_CS %>%
  filter(between(Year, starr_burn - 2, starr_burn + 2)) %>%
  mutate(period = case_when(Year < starr_burn ~ "pre",
                            Year > starr_burn ~ "post",
                            TRUE ~ "burn")) %>% 
  group_by(Plot, period) %>% 
  summarise(mean_bai = mean(measurement),
            stand = "CS")

fire_pre_post_bai_TW51 <-
  bai_TW51 %>%
  filter(between(Year, starr_burn - 2, starr_burn + 2)) %>%
  mutate(period = case_when(Year < starr_burn ~ "pre",
                            Year > starr_burn ~ "post",
                            TRUE ~ "burn")) %>% 
  group_by(Plot, period) %>% 
  summarise(mean_bai = mean(measurement),
            stand = "TW51")


fire_pre_post_bai_TB <-
  bai_TB %>%
  filter(between(Year, fs_burn - 2, fs_burn + 2)) %>%
  mutate(period = case_when(Year < fs_burn ~ "pre",
                            Year > fs_burn ~ "post",
                            TRUE ~ "burn")) %>%   
  group_by(Plot, period) %>% 
  summarise(mean_bai = mean(measurement),
            stand = "TB")


fire_pre_post_bai_CL <-
  bai_CL %>%
  filter(between(Year, fs_burn - 2, fs_burn + 2)) %>%
  mutate(period = case_when(Year < fs_burn ~ "pre",
                            Year > fs_burn ~ "post",
                            TRUE ~ "burn")) %>%   
  group_by(Plot, period) %>% 
  summarise(mean_bai = mean(measurement),
            stand = "CL")


```

Combine everything together
```{r}
fire_bai_data <- fire_pre_post_bai_CS %>% 
  bind_rows(fire_pre_post_bai_CL) %>% 
  bind_rows(fire_pre_post_bai_TB) %>% 
  bind_rows(fire_pre_post_bai_TW51) %>% 
  filter(period != "burn") %>% 
  mutate(period = factor(period, levels = c("pre", "post")))
```


```{r}
m1 <- lm(mean_bai ~ period, data = fire_bai_data)
summary(m1)
```


```{r}
library(lme4)
library(lmerTest)

mm1 <- lmer(mean_bai ~ period + (1|stand), data = fire_bai_data)
summary(mm1)

mm2 <- lmer(mean_bai ~ period + (1|stand/Plot), data = fire_bai_data)
summary(mm2)
```


...using a linear mixed model with the lme4 and lmerTest packages. Random effects included tree nested within stand.

